---
title: "Asingment 3 - Javier Fernandez"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
options(tinytex.verbose = TRUE)

```

# Information
### Assignment #3 \
### Class: Econ 613 \
### Name: Javier Fernandez \
### Date: April 26th, 2021 \

## Preliminaries -----
```{r warning=FALSE, message=FALSE}
rm(list=ls())

# Libraries
library(data.table)
library(moments)
library(stargazer)
library(tidyverse)

# Set WD
setwd("C:/Users/javie/OneDrive/Documents/GitHub/ECON613/Assingments/A3")
```


# I - Loading Datasets
```{r warning=FALSE, message=FALSE}
#Load data
population <- read.csv("population.csv")
crime <- read.csv("crime_long.csv")
```

# II - Data Manipulation
## 2.1 Calculate total crime per month and plot the time series of crime.

  * a. Total crime overall
```{r warning=FALSE, message=FALSE}
total_a <- crime %>% group_by(crime_month) %>% summarize(total_crimes=sum(crimes))
ggplot(data = total_a,aes(x=crime_month,y=total_crimes,group=1))+geom_line()
```

  * b. Total crime by district
```{r warning=FALSE, message=FALSE}
total_b <- crime %>% group_by(crime_month,district) %>% summarize(total_crimes=sum(crimes))
ggplot(data=total_b,aes(x=crime_month,y=total_crimes,
                             colour=district,group=district)) +
  geom_line()

# Mege crime by district 
crime <- left_join(crime,total_b)
```

### 2.2 Merge the two datasets by districts-units and period.
```{r warning=FALSE, message=FALSE}
crime  <- crime %>% pivot_wider(names_from=crime_type,values_from=crimes,
values_fn=sum) # As there are some rows with the same months and district and different values, 
                # we use values_fn
population <- population %>% rename(crime_month=month)
pop_crime <- left_join(crime,population)
```

### 3. Construct a panel data of unit over time with the following variables
    * Total crimes per resident 
    * Violent crimes per resident 
    * Property crimes per resident
    * Median income
    * Share of black, hispanic and white residents
```{r warning=FALSE, message=FALSE}
# Creating variables
Variables <-pop_crime %>% group_by(crime_month,district) %>% 
  summarise(tot_crim_per_res=total_crimes/tot_pop,
            viol_crim_per_res=violent/tot_pop,
            prop_crim_per_res=property/tot_pop,
            share_black=tot_black/tot_pop,
            share_hisp=tot_hisp/tot_pop,
            share_white=tot_white/tot_pop)

# Creating panel data
panel_crime <- left_join(pop_crime,Variables)
```

# III - Panel Data: Introduction
```{r warning=FALSE, message=FALSE}
#Load data
officers <- read.csv("officers.csv")

officers <- officers %>% rename(district=unit,crime_month=month)

final_panel <- left_join(officers,panel_crime)


model1 <- lm(data=final_panel,
             arrest~tenure+tot_crim_per_res+p50_inc+share_white+share_hisp+share_black) 
 
summary(model1)
```

# IV - Panel Data: More Controls
```{r}
model2 <- lm(data=final_panel,
             arrest~tenure+tot_crim_per_res+p50_inc+share_white+share_hisp+share_black+
     factor(crime_month)+factor(district))

summary(model2)
```

# V - Panel Data: Individual Fixed Effects
```{r, warning=FALSE}
### 5.1 Between estimation

between_data <- final_panel %>% as.data.table()
between_data <- between_data[,paste('avg', names(between_data)[4:22], sep="_") :=lapply(.SD,mean),
                             by=NUID,.SDcols=4:22] %>% as.data.frame()
# Renaming for presentation purposes
bet_data <- between_data %>%dplyr::select(1,2,3,23:41) %>%     rename(arrest=avg_arrest,tenure=avg_tenure,tot_crim_per_res=avg_tot_crim_per_res,
        p50_inc=avg_p50_inc,share_white = avg_share_white, share_hisp= avg_share_hisp, share_black =    avg_share_black)
bet_model <- lm(data=bet_data,arrest~tenure+tot_crim_per_res+p50_inc+
     share_white+share_hisp+share_black) 



### 5.2 Within estimation

within_data <- left_join(final_panel %>% 
                           dplyr::select(NUID,crime_month,arrest,tenure,tot_crim_per_res,
                                  p50_inc,tot_white,tot_hisp,tot_black),between_data)
within_data <- within_data %>% mutate(
  w_arrest=arrest-avg_arrest,
  w_tenure=tenure-avg_tenure,
  w_tot_crim_per_res=tot_crim_per_res-avg_tot_crim_per_res,
  w_p50_inc=p50_inc-avg_p50_inc,
  w_share_white=share_white-avg_share_white,
  w_share_hisp=share_hisp-avg_share_hisp,
  w_share_black=share_black-avg_share_black
)
# Renaming for presentation purposes
within_data <- within_data %>%  dplyr::select(1,2,42:48) %>%  rename(arrest=w_arrest,tenure=w_tenure,tot_crim_per_res=w_tot_crim_per_res,
        p50_inc=w_p50_inc,share_white = w_share_white, share_hisp= w_share_hisp, share_black =    w_share_black)

within_model <- lm(data=within_data,arrest~tenure+tot_crim_per_res+p50_inc+
     share_white+share_hisp+share_black)

### 5.3 First Difference
subset_panel<- final_panel %>% dplyr::select(NUID,crime_month,arrest,tenure,tot_crim_per_res,
                       p50_inc,share_white,share_hisp,share_black) %>% 
              mutate(NUID=as.numeric(NUID))
ordered_panel <- subset_panel[order(subset_panel$NUID,subset_panel$crime_month),]
diff_panel <- NULL

# Looking for the first observation of each individual
min_month <- ordered_panel %>% group_by(NUID) %>% summarise(crime_month=min(crime_month))
min_month <- min_month %>% mutate(drop=1)

# Computing differences
diff_panel <- ordered_panel[,3:9]-dplyr::lag(ordered_panel[,3:9])
diff_panel <- cbind(ordered_panel[,1:2],diff_panel)

# Removing the first observation of each individual  
diff_panel <- left_join(diff_panel,min_month)
diff_panel <- diff_panel %>% filter(is.na(drop))
diff_panel <- diff_panel %>% dplyr::select(-drop)

# Estimate model
firstdiff_model <- lm(data=diff_panel, arrest~tenure+tot_crim_per_res+p50_inc+
                        share_white+share_hisp+share_black)
```

```{r, results='asis'}
stargazer(bet_model,within_model,firstdiff_model,column.labels =c("Between","Within","First Difference"),title="Results",dep.var.labels = "Arrests",covariate.labels = c("Tenure","Crimes per resident","Median Income","Share of White","Share of Hispanic","Share of Black"))
```

We observe very different estimations of $\beta$ depending on the model we choose. In first place, because they represent different things but also because of the nature of the data. The only turning out significant is for the Between model. This indicates that the variation that allows for the estimation of our parameter is variation across individuals instead of variation over time. If we observe the data, variables related to the population (median income and shares by race) present only a limited series of discrete jumps, this is a large number of zeros and some large values. This is specially relevant for the first difference estimation were the only source of variation is due to irregularities in the data. Therefore, for the nature of the data observed, the between model is better than other specifications.

### 5.4 GMM estimation ----

The initial step would be to construct identification keys for individuals, months, and districts. For the latter, as its values are ordered integers, modification of the data is not necessary. 
```{r}
officers <- final_panel %>% dplyr::select(NUID) %>% unique()
officers <- officers %>% mutate(id_off=1:nrow(officers))

months <- final_panel %>% dplyr::select(crime_month) %>% unique()
months <- months %>% mutate(id_month=1:nrow(months))

gmm_panel <- left_join(final_panel,officers)
gmm_panel <- left_join(gmm_panel,months)

```

#### The Generalized Method of Moments function
``` {r}
GMM = function(param,gmm_panel){
  gmm_panel <- gmm_panel %>% drop_na()
  #Coefficients and preliminaries
  ni = max(gmm_panel$id_off)
  alpha = param[1:ni]
  beta = param[ni+1]
  gamma = param[(ni+2):(ni+6)]
  phi = param[(ni+7):(ni+31)]
  kappa = param[(ni+32):(ni+31+max(gmm_panel$id_month))]
  
  #Variables
  arrest = gmm_panel$arrest
  tenure = gmm_panel$tenure
  tot_crimes_pr = gmm_panel$tot_crim_per_res
  median_income = gmm_panel$p50_inc
  s_black= gmm_panel$share_black
  s_white= gmm_panel$share_white
  s_hisp= gmm_panel$share_hisp
  
  # Loop to compute estimates
  est=mat.or.vec(nrow(gmm_panel),1)
    for (ii in 1:nrow(gmm_panel)){
    est[ii] = alpha[gmm_panel$id_off[ii]]+beta*tenure[ii]+
      gamma[1]*tot_crimes_pr[ii] + gamma[2]*median_income[ii]+
      gamma[3]*s_black[ii]+gamma[4]*s_white[ii]+gamma[5]*s_hisp[ii]
    + phi[gmm_panel$district[ii]]+kappa[gmm_panel$id_month[ii]]
  }

  moments_est=all.moments(est,order.max=2)[-1]
  moments_arrest=all.moments(arrest,order.max=2)[-1]
  
  like = sum((moments_est - moments_arrest)^2)
  return(like)
}
```

#### GMM Estimation
Given that our sample consists of 13,029 officers along time and districts, this implies the estimation of  13,192 parametes if used the complete sample. To illustrate the results of the code, the estimation is done for the first 50,000 observations (~5% of the data).

```{r}
set.seed(123456)
subset_gmm_panel <- gmm_panel[1:50000,]  
param=runif(max(subset_gmm_panel$id_off)+6+max(subset_gmm_panel$id_month))/10000
GMM(param,subset_gmm_panel)


res  = optim(param,fn=GMM,method="BFGS",control=list(trace=6,REPORT=1,maxit=1000),
             gmm_panel=subset_gmm_panel)

```

##Results of GMM Estimation
```{r}
# Histogram of individual fixed effects 
hist(res$par[1:623], main = "Individual Fixed Effects")

# Coefficients
gmm_coef <- res$par[624:629]
names(gmm_coef) <- c("Tenure","Crimes per resident","Median Income","Share of White","Share of Hispanic","Share of Black")
gmm_coef
# Histogram of district effects
hist(res$par[630:654], main = "District Effects")

# Histogram of time effects
hist(res$par[655:761], main = "Time Effects")
```